The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-06 thesis/data.tex
full draft of appendicies, improved formatting

\section{\texorpdfstring{Satellite stellar membership with
\emph{Gaia}}{Satellite stellar membership with Gaia}}\label{satellite-stellar-membership-with-gaia}

@@

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second
exponential component is considered only when the preferred amplitude of
the second component is non-zero. The \emph{proper-motion likelihood}
quantifies the agreement of a star's motion with the dwarf galaxy's
systemic motion, accounting for observational uncertainties. The
\emph{CMD likelihood} measures the consistency of a star's \emph{Gaia}
\(G\) magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
isochrones for the dwarf. For the background likelihoods, the spatial
likelihood is a uniform distribution over the field, and the background
CMD and PM likelihoods are constructed empirically from the stars in an
annulus far from the satellite. Each likelihood is normalized as a
probability density over the respective parameter space. Appendix
\ref{sec:density_extra} discusses the details of quality cuts and the
likelihood calculation.

@@

We illustrate the likelihood-based membership selection criteria through
a progressive tightening of criteria. First, we consider a
minimally-refined sample, \textbf{all}, which only excludes stars with
poor astrometry, unreliable photometry, or inconsistent parallaxes.
Next, incorporating CMD and PM information, we use the \textbf{CMD+PM}
to test a selection method agnostic to the spatial position. This sample
includes stars where the CMD and PM combined likelihood favours
satellite membership:
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
}\end{equation} Next, our \textbf{fiducial} sample,
\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
details). While the latter is the highest-purity membership sample, we
do not use this sample for stellar density analysis due to its
incompleteness and complex selection function.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_selection.png}}
\caption[Sculptor sample selection]{The distributions of various samples
of \emph{Gaia} stars for Sculptor. In each panel, we plot light grey
points for the ``all'' sample of stars, turquoise points for ``CMD+PM''
selected stars (Eq.~\ref{eq:sel_cmd_pm}), blue squares for the
``fiducial'' sample (\(P_{\rm sat} > 0.2\)), and indigo diamonds for the
``RV members'' sample. We mark the two far-outlier stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Left:}
Tangent plane \(\xi, \eta\). The orange ellipses represents 3 and 6
half-light radii. \textbf{Top right:} Colour magnitude diagram in Gaia
\(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
black bar in the top left represents the median colour error.
\textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
\(\mu_{\alpha*}\) (corrected). The orange point marks the systemic
\citet{MV2020b} proper motion. The black cross represents the median
proper motion error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_selection.png}}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline
``velocity confirmed'' members outside a radius of \(6R_h\) with red
stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. We also
mark the location of Muñoz 1 with a pink
circle.}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/fornax_selection.png}}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. The isocrone is instead
for a 2Gyr stellar population. RV measurements are from \citet{WMO2009}.
Fornax does not show the same extended outer halo of probable members as
Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

>>>>

\section{\texorpdfstring{Satellite stellar membership with
\emph{Gaia}}{Satellite stellar membership with Gaia}}\label{sec:the_algorithm}

@@

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second
exponential component is considered only when the preferred amplitude of
the second component is non-zero. The \emph{proper-motion likelihood}
quantifies the agreement of a star's motion with the dwarf galaxy's
systemic motion, accounting for observational uncertainties. The
\emph{CMD likelihood} measures the consistency of a star's \emph{Gaia}
\(G\) magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
isochrones for the dwarf. For the background likelihoods, the spatial
likelihood is a uniform distribution over the field, and the background
CMD and PM likelihoods are constructed empirically from the stars in an
annulus far from the satellite. Each likelihood is normalized as a
probability density over the respective parameter space. Appendix
\ref{sec:density_extra} discusses the details of quality cuts and the
likelihood calculation \citep[see also,][]{MV2020a, jensen+2024}.

@@

We illustrate the likelihood-based membership selection criteria through
a progressive tightening of criteria. First, we consider a
minimally-refined sample, \textbf{all}, which only excludes stars with
poor astrometry, unreliable photometry, or inconsistent parallaxes.
Next, incorporating CMD and PM information, we use the \textbf{CMD+PM}
to test a selection method agnostic to the spatial position. This sample
includes stars where the CMD and PM combined likelihood favours
satellite membership:
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
}\end{equation} Next, our \textbf{fiducial} sample,
\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
line-of-sight velocity information (see Appendix
\ref{sec:extra_rv_models} for details). While the latter is the
highest-purity membership sample, we do not use this sample for stellar
density analysis due to its incompleteness and complex selection
function.

@@

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/scl_selection.png}
\caption[Sculptor sample selection]{The distributions of various samples
of \emph{Gaia} stars for Sculptor. In each panel, we plot light grey
points for the ``all'' sample of stars, turquoise points for ``CMD+PM''
selected stars (Eq.~\ref{eq:sel_cmd_pm}), blue squares for the
``fiducial'' sample (\(P_{\rm sat} > 0.2\)), and indigo diamonds for the
``RV members'' sample. We mark the two far-outlier stars from
\citet{sestito+2023a} with rust-outlined indigo stars. \textbf{Left:}
Tangent plane \(\xi, \eta\). The orange ellipses represents 3 and 6
half-light radii. \textbf{Top right:} Colour magnitude diagram in Gaia
\(G\) magnitude versus \(G_{\rm BP} - G_{\rm RP}\) colour. We plot a
Padova \(12\,\)Gyr, \({\rm [Fe/H] }=-1.68\) isochone in orange. The
black bar in the top left represents the median colour error.
\textbf{Bottom right:} Proper motion in declination \(\mu_\delta\) vs RA
\(\mu_{\alpha*}\) (corrected). The orange point marks the systemic
\citet{MV2020b} proper motion. The black cross represents the median
proper motion error.}\label{fig:scl_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/umi_selection.png}
\caption[Ursa Minor sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Ursa Minor. We outline
``velocity confirmed'' members outside a radius of \(6R_h\) with red
stars \citep[from][]{sestito+2023b, pace+2020, spencer+2018}. We also
mark the location of Muñoz 1 with a pink
circle.}\label{fig:umi_selection}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/fornax_selection.png}
\caption[Fornax sample selection]{Similar to
Fig.~\ref{fig:scl_selection} except for Fornax. The isocrone is instead
for a 2Gyr stellar population. RV measurements are from \citet{WMO2009}.
Fornax does not show the same extended outer halo of probable members as
Sculptor or Ursa Minor despite having many more
stars.}\label{fig:fornax_selection}
\end{figure}

### 2025-10-06 thesis/discussion.tex
full draft of appendicies, improved formatting

Tidal effects have been long invoked to explain features of dwarf
galaxies e.g.~Section \ref{sec:scl_umi_obs_tides},
\citep{mayer+2001a, tsujimoto+shigeyama2002} \textbf{Check this
citation}. Yet, growing work suggests tides may not be as commonplace as
suspected. For example, \citet{read+2006} suggest that the absence of
rising radial velocity dispersion profiles in dwarfs implies a lack of
tidal disruption. \citet{penarrubia+2009}, introducing the break radius
as a diagnostic (Eq.~\ref{eq:r_break}), showed that most satellite
orbits were inconsistent with observable tidal effects. Recently,
\citet{pace+erkal+li2022} use a Jacobi radius-based criterion
(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. With the
exception of the Sagittarius dSph, these results suggest tides have mild
influence on the stellar components of classical dwarfs.

>>>>

Tidal effects have been long invoked to explain features of dwarf
galaxies \citep[e.g.,
Section~\ref{sec:scl_umi_obs_tides},][]{mayer+2001a, tsujimoto+shigeyama2002}
\textbf{Check this citation}. Yet, growing work suggests tides may not
be as commonplace as suspected. For example, \citet{read+2006} suggest
that the absence of rising radial velocity dispersion profiles in dwarfs
implies a lack of tidal disruption. \citet{penarrubia+2009}, introducing
the break radius as a diagnostic (Eq.~\ref{eq:r_break}), showed that
most satellite orbits were inconsistent with observable tidal effects.
Recently, \citet{pace+erkal+li2022} use a Jacobi radius-based criterion
(Eq.~\ref{eq:r_jacobi}) to reach a similar conclusion. With the
exception of the Sagittarius dSph, these results suggest tides have mild
influence on the stellar components of classical dwarfs.
